Methods paper for erosion

### 2017-11-11 docs/methods/erosion_methods.tex
Worked on methods paper, theta-L evolution with RK2.

For simplicity, we introduce the variables
\begin{align}
\label{eq:M}
& M = - \mean{\thalpha \Vn} \\
\label{eq:N}
& N = \frac{1}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right) \\
& \elfun = \frac{1}{L^{2} \log \left(2\pi/L \right)}
\end{align} 
Then Eq.~\eqref{eqn:thetaevo0} can be rewritten as
\begin{equation}
\label{eqn:thetaevo1}
\pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + N
\end{equation}
Note that the $\elfun$ depends only on $L(t)$, and so the highest order
derivative in Eq.~\eqref{eqn:thetaevo1} is {\em linear} in the variable $\alpha$. This is the key to the {\thL} method, and allows us to use very stable, implicit methods for the stiff term. We will evolve $\theta$ in spectral space, and so introduce the transforms
\begin{align}
& \theta(\alpha,t) = \sum_{k = -\infty}^{\infty} \thhat_k e^{2\pi i k \alpha} \\
& N(\alpha,t)  = \sum_{k = -\infty}^{\infty} \hat{N}_k e^{2\pi i k \alpha}
\end{align}
\todo[inline]{Earlier, I assumed the geometry is $2\pi$ periodic.  Is it
possible to use this convention cleanly for the presentation?  This
allows us to avoid the long superscripts $2\pi i k \alpha$}
Equation~\eqref{eqn:thetaevo1} gives a set of ODEs for the coefficients
\begin{equation}
\tderiv{\thhat_k}{t} +  4 \eps \pi^2 k^2  \elfun \thhat_k = \hat{N}_k
\end{equation}
We introduce the integrating factor
\begin{equation}
\label{eq:mu}
\mu_k = \exp \left( 4 \eps \pi^2 k^2 \int \elfun \, dt \right)
\end{equation}
Then the evolution equations for $L$ and $\theta$ can be recast as the coupled system of ODEs
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = M \\
\label{eqn:thetaevo2}
& \tderiv{}{t}\left( \mu_k \theta_k \right) = \mu_k \hat{N}_k
\end{align}
We solve system~\eqref{eqn:Levo2}--\eqref{eqn:thetaevo2} for the
variables $L$ and $\{ \mu_k \theta_k \}$ with a 2nd-order Runge-Kutta
method. At each stage, we must recover $\theta(\alpha,t)$ from  $\{
  \mu_k \theta_k \}$ in order to compute the shear stress as needed in
  Eqs.~\eqref{eq:M} and~\eqref{eq:N} for $M$ and $N$. For this, we
  compute $\mu_k$ from Eq.~\eqref{eq:mu} using the trapezoid rule, then recover $\theta(\alpha, t)$ via a Gaussian filter....

>>>>

For simplicity, we introduce the variables
\begin{align}
\label{eq:M}
& M = - \mean{\thalpha \Vn} \\
\label{eq:N}
& N = \frac{1}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right) \\
\label{eq:elfun}
& \elfun = \frac{1}{L^{2} \log \left(2\pi/L \right)}
\end{align} 
Then Eq.~\eqref{eqn:thetaevo0} can be rewritten as
\begin{equation}
\label{eqn:thetaevo1}
\pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + N
\end{equation}
Note that the $\elfun$ depends only on $L(t)$, and so the highest order
derivative in Eq.~\eqref{eqn:thetaevo1} is {\em linear} in the variable $\alpha$. This is the key to the {\thL} method, and allows us to use very stable, implicit methods for the stiff term. We will evolve $\theta$ in spectral space, and so introduce the transforms
\begin{align}
& \theta(\alpha,t) = \sum_{k = -\infty}^{\infty} \thhat_k e^{2\pi i k \alpha} \\
& N(\alpha,t)  = \sum_{k = -\infty}^{\infty} \hat{N}_k e^{2\pi i k \alpha}
\end{align}
\todo[inline]{Earlier, I assumed the geometry is $2\pi$ periodic.  Is it
possible to use this convention cleanly for the presentation?  This
allows us to avoid the long superscripts $2\pi i k \alpha$}
Equation~\eqref{eqn:thetaevo1} gives a set of ODEs for the coefficients
\begin{equation}
\tderiv{\thhat_k}{t} +  4 \eps \pi^2 k^2  \elfun \thhat_k = \hat{N}_k
\end{equation}
We introduce the integrating factor
\begin{equation}
\label{eq:mu}
\mu_k = \exp \left( 4 \eps \pi^2 k^2 \int \elfun \, dt \right)
\end{equation}
Then the evolution equations for $L$ and $\theta$ can be recast as the coupled system of ODEs
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = M \\
\label{eqn:thetaevo2}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \hat{N}_k
\end{align}
We solve system \eqref{eqn:Levo2}--\eqref{eqn:thetaevo2} with a combination of an explicit, second-order Runge-Kutta method (RK2) and trapezoid rule to compute the integrating factor, $\mu_k$. There is not a unique second-order Runge-Kutta method, and we choose the {\em midpoint} version to maximize the smoothing effect of the curvature term.

At the first stage of RK2 (the half step), we discretize Eqs.~\eqref{eqn:Levo2}--\eqref{eqn:thetaevo2} as
\begin{align}
\label{eq:L12}
& \frac{L^{(n+1/2)} - L^{(n)}}{\Dt/2} = M^{(n)} \\
& \frac{ \mu_k^{(n+1/2)} \thhat_k^{(n+1/2)} - \thhat_k^{(n)}}{\Dt/2} = \hat{N}_k^{(n)}
\end{align}
where the superscript indicates the time step. For simplicity, we have chosen the constant of integration in the exponent of Eq.~\eqref{eq:mu} so that $\mu_k^{(n)} = 1$. At the next stage of RK2, we must compute $M^{(n+1/2)}$ and $\hat{N}_k^{(n+1/2)}$, which requires a call to the Stokes solver. Hence, we must compute compute $\mu_k^{(n+1/2})$, so that $\thhat_k^{(n+1/2)}$ can be recovered and the geometry at the mid-step can be computed. To compute $\mu_k^{(n+1/2)}$, we use the trapezoid rule
\begin{equation}
\mu_k^{(n+1/2)} = \exp \left( \eps \pi^2 k^2 \Dt \left( \elfun^{(n)} + \elfun^{(n+1/2)} \right) \right)
\end{equation}
Notice in Eq.~\eqref{eq:elfun}, that $\elfun$ depends only on $L(t)$. Hence, we just need to update $L^{(n+1/2)}$ first via Eq.~\eqref{eq:L12}, and we have what we need to compute $\mu_k^{(n+1/2)}$. It is instructive to recast as
\begin{equation}
\thhat_k^{(n+1/2)} = \left( \thhat_k^{(n)} + \frac{1}{2} \Dt \, \hat{N}_k^{(n)} \right)
\exp \left( - \eps \pi^2 k^2 \Dt \left( \elfun^{(n)} + \elfun^{(n+1/2)} \right) \right)
\end{equation}
so that the smoothing effect of the exponential is obvious. With $\thhat_k^{(n+1/2)}$ computed, it is now possible to compute $M^{(n+1/2)}$ and $\hat{N}_k^{(n+1/2)}$.

Next, at the second stage of RK2, we discretize as
\begin{align}
& \frac{L^{(n+1)} - L^{(n)}}{\Dt} = M^{(n+1/2)} \\
& \frac{ \mu_k^{(n+1)} \thhat_k^{(n+1)} - \thhat_k^{(n)}}{\Dt} = \hat{N}_k^{(n+1/2)}
\end{align}
We again compute the integrating factor, $\mu_k^{(n+1)}$ via the trapezoid rule, giving
\begin{equation}
\thhat_k^{(n+1)} = \left( \thhat_k^{(n)} + \Dt \, \hat{N}_k^{(n+1/2)} \right)
\exp \left( - 2 \eps \pi^2 k^2 \Dt \left( \elfun^{(n)} + \elfun^{(n+1)} \right) \right)
\end{equation}
